MSCA 31006: Final Project - AQI Forecasting

### 2023-08-13 MSCA_31006_Time_Series_Final_Project.Rmd
Add files via upload

#### Step 3: Extract weekly values for time series and plot the series

```{r, tidy=TRUE}
# Create weekly data
pm2_5_df <- as.data.frame(pm2_5_ts)
pm2_5_df$Date <- mdy(pm2_5_data$Date)
colnames(pm2_5_df) <- c("PM2.5.AQI.Value", "Date")

# Convert to xts format for weekly conversion
pm2_5_xts <- as.xts(pm2_5_df, order.by=pm2_5_df$Date)
pm2_5_xts <- pm2_5_xts[,-2]
pm2_5_weekly <- apply.weekly(pm2_5_xts, mean)
#pm2_5_weekly

pm2_5_weekly_ts <- ts(pm2_5_weekly["19990103/20230811"], start = c(1999,1),frequency = 52.18)
plot(pm2_5_weekly_ts)

@@

#### Step 4: Extract monthly values for time series and plot the series

```{r, tidy=TRUE}
# Create monthly data
pm2_5_monthly <- apply.monthly(pm2_5_xts, mean)
#pm2_5_monthly

@@

#### Step 5: Understand time series variance

plot(pm2_5_weekly_ts)

# Strong seasonality, Strong cyclic, light downward trend, variance is reducing

# Let's see how seasonality looks over time and is the variance changing
pm2_5_df$Month <- month(pm2_5_df$Date)
pm2_5_df$Year <- year(pm2_5_df$Date)

# We can see that over the years there is downward trend and variance is decreasing
# except 2020 when we see a high peak likely caused by wildfires
# Link: https://en.wikipedia.org/wiki/Lake_Fire_(2020)

#### Step 6: Split data into train and test - Weekly

# Split the data into a training and test dataset
train <- window(pm2_5_weekly_ts, start = c(1999,1), end=c(2020,52))
test <- window(pm2_5_weekly_ts, start = c(2021,1), end=c(2023,7))

#### Step 7: Decompose the time series plot - Weekly

# Looking at spectral analysis
periodogram(train, log = "no", plot=TRUE, 
            ylab = "Periodogram",
            xlab = "Frequency",
            lwd=2, xlim = c(0, 0.06))

# Decompose the series
plot(decompose(train, type="multiplicative"))

# Understand the seasonality and remove it to see the trend
train_diff <- diff(train, lag = 52)
autoplot(train_diff, ylab="Train Seasonal Differencing")

# Let's look if the series is stationary?
kpss.test(train_diff)
# The reported p-value is 0.1, which is > 0.05, and would suggest that we fail 
# to reject the null hypothesis of level stationarity (conclusion: stationary)

adf.test(train_diff)
# The reported p-value of 0.01 (<0.05) so we do reject the null hypothesis that 
# the time series is non-stationary (conclusion: stationary)

acf(train_diff, main = "ACF of Seasonal Differencing Time Series")
pacf(train_diff, main = "PACF of Seasonal Differencing Time Series")

#### Step 8: Benchmark using snaive model - Weekly

# Forecast the seasonal naïve for (2021-01 to 2023-07)
forecast_snaive <- snaive(train, h=110)
# Plot the forecasts for snaive model
plot(forecast_snaive, main = "PM 2.5 AQI Forecast - SNaive",
         xlab = "Week", ylab = "PM 2.5 AQI")
lines(test)

# Compare the forecast with the actual test data by calculating MAPE and MSE
# Mean Absolute Percentage Error (MAPE)
mape_snaive <- mean(abs((test - forecast_snaive$mean)/test))
mape_snaive

# Mean Squared Error (MSE)
mse_snaive <- mean((test - forecast_snaive$mean)^2)
mse_snaive

#### Step 9: Split data into train and test - Monthly

# Split the data into a training and test dataset
train_monthly <- window(pm2_5_monthly_ts, start = c(1999,1), end=c(2020,12))
test_monthly <- window(pm2_5_monthly_ts, start = c(2021,1), end=c(2023,7))

#### Step 10: Decompose the time series plot - Monthly

@@

#### Step 11: Benchmark using snaive model - Monthly

@@

>>>>

```
#### Step 3: Prepare data for Weekly & Monthly extraction and EDA

```{r, tidy=TRUE}
# Create data with Dates
pm2_5_df <- as.data.frame(pm2_5_ts)
pm2_5_df$Date <- mdy(pm2_5_data$Date)
colnames(pm2_5_df) <- c("PM2.5.AQI.Value", "Date")

# Convert to xts format for weekly & monthly
pm2_5_xts <- as.xts(pm2_5_df, order.by=pm2_5_df$Date)
pm2_5_xts <- pm2_5_xts[,-2]
```

#### Step 4: Understand PM2.5 AQI variance across years

```{r, tidy=TRUE}
# Let's see how seasonality looks over time and is the variance changing
pm2_5_df$Month <- month(pm2_5_df$Date)
pm2_5_df$Year <- year(pm2_5_df$Date)

# We can see that over the years there is downward trend and variance is decreasing
# except 2020 when we see a high peak likely caused by wildfires
# Link: https://en.wikipedia.org/wiki/Lake_Fire_(2020)
```

#### Step 5a: Extract values for time series and plot the series - Weekly

```{r, tidy=TRUE}
pm2_5_weekly <- apply.weekly(pm2_5_xts, mean)
pm2_5_weekly_ts <- ts(pm2_5_weekly["19990103/20230811"], 
                      start = c(1999,1),
                      frequency = 52.18)
plot(pm2_5_weekly_ts)
# Strong seasonality, Strong cyclic, light downward trend, variance is reducing

@@

#### Step 6a: Split data into train and test - Weekly

```{r, tidy=TRUE}
# Split the data into a training and test dataset
train_weekly <- window(pm2_5_weekly_ts, start = c(1999,1), end=c(2020,52))
test_weekly <- window(pm2_5_weekly_ts, start = c(2021,1))
```

#### Step 7a: Decompose the time series plot - Weekly

```{r, tidy=TRUE}
# Looking at spectral analysis
periodogram(train_weekly, log = "no", plot=TRUE, 
            ylab = "Periodogram",
            xlab = "Frequency",
            lwd=2, xlim = c(0, 0.06))

# Decompose the series
plot(decompose(train_weekly, type="multiplicative"))

# Understand the seasonality and remove it to see the trend
train_weekly_diff <- diff(train_weekly, lag = 52)
autoplot(train_weekly_diff, ylab="Train Seasonal Differencing - Weekly Data")

# Let's look if the series is stationary?
kpss.test(train_weekly_diff)
# The reported p-value is 0.1, which is > 0.05, and would suggest that we fail 
# to reject the null hypothesis of level stationarity (conclusion: stationary)

adf.test(train_weekly_diff)
# The reported p-value of 0.01 (<0.05) so we do reject the null hypothesis that 
# the time series is non-stationary (conclusion: stationary)

acf(train_weekly_diff, main = "ACF of Seasonal Differencing Time Series - Weekly Data")
pacf(train_weekly_diff, main = "PACF of Seasonal Differencing Time Series - Weekly Data")

#### Step 8a: Benchmark using snaive model - Weekly

```{r, tidy=TRUE}
# Forecast the seasonal naïve for (2021-01 to 2023-07)
forecast_snaive_weekly <- snaive(train_weekly, h=110)
# Plot the forecasts for snaive model
plot(forecast_snaive_weekly, main = "PM 2.5 AQI Forecast - SNaive (Weekly)",
         xlab = "Week", ylab = "PM 2.5 AQI")
lines(test_weekly)

# Compare the forecast with the actual test data by calculating MAPE and MSE
# Mean Absolute Percentage Error (MAPE)
mape_snaive_weekly <- mean(abs((test_weekly - forecast_snaive_weekly$mean)/test_weekly))
mape_snaive_weekly

# Mean Absolute Error (MAE)
mae_snaive_weekly <- mean(abs((test_weekly - forecast_snaive_weekly$mean)))
mae_snaive_weekly

# Mean Squared Error (MSE)
mse_snaive_weekly <- mean((test_weekly - forecast_snaive_weekly$mean)^2)
mse_snaive_weekly

# Evaluate the residuals
checkresiduals(forecast_snaive_weekly)

```
#### Step 5b: Extract values for time series and plot the series - Monthly

```{r, tidy=TRUE}
# Create monthly data
pm2_5_monthly <- apply.monthly(pm2_5_xts, mean)

@@

#### Step 6b: Split data into train and test - Monthly

```{r, tidy=TRUE}
# Split the data into a training and test dataset
train_monthly <- window(pm2_5_monthly_ts, start = c(1999,1), end=c(2020,12))
test_monthly <- window(pm2_5_monthly_ts, start = c(2021,1))
```

#### Step 7b: Decompose the time series plot - Monthly

@@

#### Step 8b: Benchmark using snaive model - Monthly

@@

# Evaluate the residuals
checkresiduals(forecast_snaive_monthly)
